Towards Comparable Active Learning

### 2024-05-30 doc/al_benchmark_paper.tex
paper

\documentclass[]{article}
%\usepackage[a4paper, total={6.5in, 8.5in}]{geometry}
\usepackage{color}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage{neurips_data_2023}

@@

\title{Towards Comparable Active Learning}
\author{%
	Anonymous Authors
	%Thorben Werner 
	%\thanks{Institute of Computer Science - Information Systems and Machine Learning Lab (ISMLL)} 
	%\\
	%University of Hildesheim\\
	%Universitätsplatz 1\, 31141 Hildesheim \\
	%\texttt{werner@ismll.de} \\
	% examples of more authors
	%\And
	%Johannes Burchert$^*$ \\
	%University of Hildesheim\\
	%Universitätsplatz 1, 31141 Hildesheim \\
	%\texttt{burchert@ismll.de} \\
	%\AND
	%Prof. Lars Schmidt-Thieme$^*$ \\
	%University of Hildesheim\\
	%Universitätsplatz 1, 31141 Hildesheim \\
	%\texttt{schmidt-thieme@ismll.uni-hildesheim.de}
}

@@

@@

\begin{abstract}
	Active Learning has received significant attention in the field of machine learning for its potential in selecting the most informative samples for labeling, thereby reducing data annotation costs. However, the lack of reproducibility of performance gains reported in recent literature has created a chaotic landscape in Active Learning research. 
	This paper addresses these issues of inconsistent results in active learning (AL) literature.
	%Previous papers are constantly reporting significant performance improvements, while subsequent literature fails to reproduce those results. This inconsistency leads to a chaotic landscape of AL algorithms.
	To the best of our knowledge, we propose the first AL benchmark that tests algorithms in 3 major domains: Tabular, Image and Text.
	Furthermore, we highlight overlooked problems for reproducing AL experiments that can lead to unfair comparisons and increased variance in the results.
	To tackle these challenges, we propose an experimental protocol to accurately control the experiments.
	We report empirical results for 6 widely used algorithms on 7 datasets and aggregate them into a domain-specific ranking of AL algorithms.
\end{abstract}

@@

Since Deep NNs usually require large amounts of labeled training data, Active Learning (AL) can be used to instead to select the most informative samples out of a large pool of unlabeled data, so that only these samples need to be labeled.

@@

%On top of providing a principled way to labeled unlabeled datasets, active learning is one of the two major approaches besides semi-supervised learning to make deep learning models more data efficient by requiring only a limited set of manually labeled data.

@@

In the last decade many different algorithms for AL have been proposed.
Even though, almost every method has reported lifts all it's predecessors 
\footnote{out of all considered algorithms for this paper, only BALD \cite{gal2017deep} did not claim a new SOTA performance in their result section}
\let\thefootnote\relax\footnotetext{Code available at: \url{https://github.com/wernerth94/comparable-active-learning}}
, AL research faces two central difficulties:

@@

(ii) the reported results prove to be very difficult to reproduce.
While multiple benchmark suites have been proposed to solve (i), to the best of our knowledge, we are the first to compare AL algorithms on all 3 data domains of vector, image and text.
Regarding (ii), \cite{zhou2021towards} has pointed out severe inconsistencies in results of AL papers in recent years. 

@@

The result of these inconsistencies is a chaotic landscape of AL algorithms where every paper claims to archive state-of-the-art performance
%by significantly outperforming everyone else
, while the vast majority of results proves to be non-reproducible. \\
In this work we propose an evaluation protocol that was designed to cope with the high variance in the performances of AL algorithms as well as being fully controllable regardless of the combination of dataset, model and AL algorithm.\\
We focus our work on single-sample pool-based AL where a pool of unlabeled samples is fixed at the start of each experiment and samples are chosen sequentially.
Specifically, we are not experimenting on so-called batch Active Learning, where at each iteration multiple unlabeled samples are chosen at the same time.
Even though most proposed algorithms (and benchmarks) are batched, they do not have a principled advantage over single-sample AL except speed of computation.
The problem of optimizing a portfolio of unlabeled samples in each iteration is more complicated to solve and the algorithms have systematically less information per sample to work with.
A performance comparison of batch AL and single-sample AL can be found in Fig. \ref{fig:ablation_bald}.
We can see that for two established algorithms BALD \cite{krizhevsky2009learning} and BADGE \cite{ashdeep} Batch AL can at most perform on-par with single-sample AL, independent of wether algorithm was originally designed for batch AL or not. \\

@@

@@

	\begin{tabular}{l | c c c c c}
		Paper & Sampling & \#Datasets & \#Domains & \#Algorithms & Oracle \\

@@

		Beck et al. \cite{beck2021effective} & batch & 4 & 1 & 7 & - \\
		Hu et al. \cite{hu2021towards} & batch & 5 & 2 & 13 & - \\
		Li et al. \cite{li2022empirical} & batch & 5 & 1 & 13 & - \\
		Zhou et al. \cite{zhou2021towards} & batch & 3 & 2 & 2 & \checkmark \\
		\textbf{Ours} & single & 7 & 3 & 6 & \checkmark 

@@

	\caption{Comparison of our benchmark with the existing literature. Oracle curves serve as an approximation of the best possible AL algorithm.}

@@

\subsection*{Contributions}
\begin{enumerate}
	\item Evaluation of Active Learning algorithms on datasets from 3 different domains
	\item Two novel synthetic dataset that highlight principled shortcomings of existing AL algorithms
	\item Novel experimental protocol for controlling AL experiments to allow full reproducibility
	\item Simple algorithm for an Oracle that can be constructed greedily and does not rely on search
\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Problem Description}
Given two spaces $\mathcal{X}, \mathcal{Y}$, $n=l+u$ data points with $l \in \mathbb{N}$ labeled examples $\mathcal{L} = \{(x_1, y_1),\ldots, (x_l,y_l)\}$, $u \in \mathbb{N}$ unlabeled examples $\mathcal{U} = \{x_{l+1},\ldots,x_{n}\}$, a budget $\mathbb{N} \ni b \le u$ and an annotator $A: \mathbb{R}^\mathcal{X} \to \mathbb{R}^\mathcal{Y}$ that can label $x$. 
We call $x \in \mathcal{X}$, $y \in \mathcal{Y}$ predictors and labels respectively where $(x,y)$ are drawn from an unknown distribution $\rho$. 
Find an acquisition function $\Omega: \U^{(i)},\LL^{(i)} \mapsto x^{(i)}$ that iteratively selects the next unlabeled point $x^{(i)} \in \U^{(i)}$ for labeling
\begin{align*}
	\LL^{(i+1)} &\gets \LL^{(i)} \cup \{\left(x^{(i)}, A(x^{(i)})\right)\} \\
	\U^{(i+1)} &\gets \U^{(i)} \setminus x^{(i)} %, i \in 1{:}B
\end{align*}
so that the expected loss $\ell: \mathbb{R}^\mathcal{Y} \times \mathbb{R}^\mathcal{Y} \to \mathbb{R}$ of the machine learning algorithm $\hat y$ after $B$ iterations is minimal: 
$$\min \quad \mathbb{E}_{(x,y) \sim \rho} \ell(y, \hat{y}(\LL^{(B)}))$$
This formulation is a special case of the problem formulation in Appendix \ref{app:problem_formulation}.

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Related Work}
\begin{figure}
	\centering
	\includegraphics[width=0.49\linewidth]{img/ablation_splice_bald}
	\includegraphics[width=0.49\linewidth]{img/ablation_splice_badge}
	\caption{Performance comparison of BALD \cite{gal2017deep} with it's batch AL adaptation \cite{kirsch2019batchbald} and BADGE \cite{ashdeep} with it's single-sample AL adaptation ($k=1$ in \cite{ashdeep} Alg. 1).
	In either scenario, the batched algorithm can at most perform on-par with it's single-sample analog, independent of wether it was originally designed for batch AL.}
	\label{fig:ablation_bald}
\end{figure}
Multiple benchmark suites have already been proposed for Active Learning:
The authors of \cite{beck2021effective} and \cite{li2022empirical} both focus exclusively on Batch AL in the image domain.
While \cite{beck2021effective} discuss a new metric to measure AL performance, which they call ``Label Efficiency'' and provide experiments on many common configurations of data preparation, model training and other hyperparameters, \cite{li2022empirical} focuses on combined approaches of AL and semi-supervised learning.
The authors of \cite{hu2021towards} study models that are trained with AL techniques in the image and text domain.
They test for several different properties of the models including robustness, response to compression techniques and final performance.
\cite{zhou2021towards} proposed an oracle algorithm for AL that uses Simulated Annealing search to approximate a solution for the optimal subset of labeled data.
Additionally, they study the generalization behavior of subsets of labeled data in the text an image domain.
The employed AL algorithms for our experiments are introduced in Section \ref{sec:sampling_strategies}.

@@

@@

\subsection{Evaluation}\label{sec:evaluation}
Following \cite{zhou2021towards}, the quality of an AL algorithm is evaluated by an ``anytime" protocol that incorporates classification performance at every iteration, as opposed to evaluating final performance after the budget is exhausted.
We employ the normalized area under the accuracy curve (AUC):
\begin{equation}\label{eq:auc}
	\operatorname{AUC}(\mathcal{D}_{test}, \hat y, B) := \frac{1}{B} \sum_{i=1}^{B} \operatorname{Acc}(\mathcal{D}_{test}, \hat y_i)
\end{equation}
where $\hat y_i$ is the (re-)trained classification model after the i-\textit{th} iteration.
To mimic the leave-one-out protocol for cross-validation we will restart each experiment multiple times.
Each restart will retain the train/test split (often given by the dataset itself), but introduces a new validation split.
The AUC incorporates performance in early stages (low budget) as well as capabilities to push the classifier in later stages (high budget).
AL algorithms have to perform well in both scenarios. 
Since AL performance inhibits high variance and is prone to outliers, we propose to aggregate AUC values with their median instead of mean. \\ [1mm]
Since the AUC depends on the chosen budget, we need a general rule on how to set this hyperparameter that does not inherently benefit a subset of algorithms.
In this work, we choose the budget per dataset to be the first point at which any algorithm (except oracle) manages to reach a percentage of the upper bound performance measured on the full dataset.
Even though we would like to propose a single percentage value for all datasets, we found that different data modalities and use-cases need different percentages to produce sensible budgets
\footnote{Since we only consider single-sample AL, increasing the budget comes at a high computational cost. Therefore we could not freely increase the budget above 2000.}.
We propose the following values: \textbf{Tabular}: 99\%, \textbf{Image}: 90\%, \textbf{Text}: 95\%, \textbf{Embedded Data}: 90\%. \\ [1mm]
Additionally, we provide evidence in Fig. \ref{fig:restarts} that previous works have not evaluated their experiments with a sufficient number of restarts.
To create Fig. \ref{fig:restarts} we used 50 restarts from the margin/random acquisition function on the Splice dataset.
From these 50 runs we uniformly sampled subsets of runs and calculated the median AUC for this subset.
One of these median AUC values corresponds to one cross-validated experiment sampled from the distribution of experiments that are restarted exactly this many times.
To create one slice in Fig. \ref{fig:restarts}, we drew 50 samples from this distribution.
Each box-plot represents the variance of an evaluation if conducted with the respective number of restarts.
We can observe that low repetitions ($<10$) provide an uncertain evaluation where lucky and unlucky draws of the same experiment give drastically different median AUC values.

@@

	\includegraphics[width=\linewidth]{img/ablation_restarts}
	\caption{Random draws from an experimental distribution on the Splice dataset with different numbers of repetitions. Each point on the Y-axis represents a cross-validated result that could have been reported in a paper. This analysis shows the drastic differences in performance one could observe even when repeating an experiment 2-10 times.}
	\label{fig:restarts}

@@

To reliably arrive at the true median AUC, we propose to repeat every experiment 50 times, as only $>42$ repetitions don't produce outliers anymore (as indicated by the rightmost columns in Fig \ref{fig:restarts}).
One way to reduce the number of necessary repetitions would be to reduce the amount of variance in the experiment through specialized seeding (discussed in the next section).
We ultimately decided in favor of high variance and high number of repetitions as the high variance accurately reflects real world applications of AL.
%Fixing for example the classifier initialization to random draws from a constant seed does not have a corresponding use-case in the real world.

>>>>

\documentclass[]{article}
%\usepackage[a4paper, total={6.5in, 8.5in}]{geometry}

%\usepackage{PRIMEarxiv}

\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage[normalem]{ulem}
\useunder{\uline}{\ul}{}
\usepackage{makecell}
\usepackage{neurips_data_2024}

\graphicspath{/}

@@

%\pagestyle{fancy}
%\thispagestyle{empty}
%\rhead{ \textit{ }} 
%\fancyhead[LO]{Running Title for Header}

\title{Towards Comparable Active Learning}
\author{%
	Anonymous Authors
%	Thorben Werner 
%	\thanks{Institute of Computer Science - Information Systems and Machine Learning Lab (ISMLL)} 
%	\\
%	University of Hildesheim\\
%	Universitätsplatz 1\, 31141 Hildesheim \\
%	\texttt{werner@ismll.de} \\
%	\And
%	Johannes Burchert$^*$ \\
%	University of Hildesheim\\
%	Universitätsplatz 1, 31141 Hildesheim \\
%	\texttt{burchert@ismll.de} \\
%	\And
%	Maximilian Stubbemann$^*$ \\
%	University of Hildesheim\\
%	Universitätsplatz 1, 31141 Hildesheim \\
%	\texttt{stubbemann@ismll.de} \\
%	\AND
%	Lars Schmidt-Thieme$^*$ \\
%	University of Hildesheim\\
%	Universitätsplatz 1, 31141 Hildesheim \\
%	\texttt{schmidt-thieme@ismll.uni-hildesheim.de}
}

@@

% Review point
% - Restarting process seems to be hard to understand (probably in terms of the seeding)
% - Justification for datasets and algorithms

@@

\begin{abstract}
Active Learning has received significant attention in the field of machine learning. Its goal is to select the most informative samples for labeling, thereby reducing data annotation costs. 
However, it has been brought to attention multiple times that reported lifts from literature generalize poorly and display high variance, leading to an inconclusive landscape in Active Learning research.
Based on recent insights for reliable evaluation for Active Learning, this work extends experimentation from the commonly used image domain to a wide spectrum of domains.
%This work follows up on recent benchmarking suites for Active Learning, applying the proposed solutions and evaluation protocols to a wide spectrum of domains.
Additionally, we provide an analysis of how many repetitions an Active Learning experiment needs in order to derive conclusive results and propose that previous benchmarks have not met the necessary number of repetitions.
%evidence that existing benchmarks might not have repeated their experiments often enough.
To the best of our knowledge, we propose the first AL benchmark that applies state-of-the-art evaluation on algorithms in 3 major domains: Tabular, Image, and Text as well as synthetic data.
We report empirical results for 10 widely used algorithms on 7 real-world and 2 synthetic datasets and aggregate them into domain-specific and overall rankings of AL algorithms.
\end{abstract}

@@

Since Deep NNs usually require large amounts of labeled training data, Active Learning (AL) can be used to instead select the most informative samples out of a large pool of unlabeled data, so that only these samples need to be labeled.

@@

On top of providing a principled way to label unlabeled datasets, active learning is one of the two major approaches besides semi-supervised learning to make deep learning models more data efficient by requiring only a limited set of manually labeled data.

@@

In the last decade, many different algorithms for AL have been proposed.
Even though, almost every method has reported lifts over all its predecessors,
\footnote{Out of all considered algorithms for this paper, only BALD \cite{gal2017deep} did not claim a new SOTA performance in their result section.}
AL research faces four central difficulties:
{\let\thefootnote\relax\footnotetext{Code available at: anonymous}}
%\let\thefootnote\relax\footnotetext{Code available at: \url{https://github.com/wernerth94/comparable-active-learning}}

@@

(ii) generalize poorly across different domains,
(iii) the reported results can be subject to very high variance across restarts and
(iv) are not always compared against important baselines like margin sampling \cite{wang2014new}.
While multiple benchmark suites have been proposed to solve (i), to the best of our knowledge, we are the first to report results on all 3 data domains of tabular, image and text.
Additionally, we provide synthetic datasets to highlight principled shortcoming of existing AL algorithms.
Regarding (ii) and (iii), \cite{zhou2021towards} has pointed out severe inconsistencies in results of AL papers in recent years. 

@@

These outlined issues lead to an inconclusive landscape of AL algorithms, where the vast majority of reported lifts are neither statistically significant, nor prove to be generalizable.
This makes it very hard to identify the best AL algorithm, or even identifying state-of-the-art algorithms.
In this work we propose an evaluation protocol that was designed to handle the high variance in the performances of AL algorithms as well as being fully controllable regardless of the combination of dataset, model and AL algorithm. 
We base our work largely on \cite{ji2023randomness}, following their guidelines for a reliable evaluation of AL algorithms, while extending the number of evaluated data domains from 1 to 5. \\
We focus on pool-based AL where a pool of unlabeled samples is fixed at the start of each experiment and one or more samples are chosen sequentially.
In addition to the default scenario of selecting a batch of samples every iteration we incorporate the single sample case into our benchmark.
Batched algorithms (and benchmarks) do not have a principled advantage over single-sample AL except for speed of computation.
The problem of optimizing a portfolio of unlabeled samples in each iteration is more complicated to solve and the algorithms have systematically less information per sample to work with leading to a generally worse performance, that impacts some algorithms more than others. 
%In Fig. \ref{fig:batch_vs_single} we present a comparison of Badge \cite{ashdeep} and Coreset\cite{sener2017active}, in the batch and single-sample setting.
%Even though both algorithms were designed for batch sampling, both perform better in the single-sample setting.
We propose single-sample AL as an important tool to identify the best acquisition function, rather than the best combination of acquisition function and diversity regularization. \\
%    \includegraphics[width=\linewidth]{img/ablation_batch_single}
%    \caption{Comparison of single-sample vs. batch selection for two acquisition functions that were specifically designed for batch selection. Even when designed for batch selection, acquisition functions tend to perform better in the single sample setting.}
%    \label{fig:batch_vs_single}

@@

We offer our datasets in two versions - normal and encoded.
An encoded dataset was pre-encoded by an self-supervised encoder model, providing a different use case for active learning.
Our synthetic and text datasets (which use word embeddings in the normal setting) are exempt from this, bringing our total dataset count to 14 across 5 domains (Tabular, Image, Text, Synthetic, Encoded).

@@

        \caption{Comparison of our benchmark with the existing literature. Oracle curves serve as an approximation of the best possible AL algorithm. Including the encoded versions of our datasets we reach 14 datasets. "Semi" indicates whether the paper is employing any form of self- or semi-supervised learning. A "-" for repetitions means that we could not determine how often each experiment is repeated in the respective framework.}
\begin{tabular}{lccccllllcl}

@@

Paper & Sampling & \#Data & \#Alg & \multicolumn{1}{l}{Img} & Txt & Tab & Synth & Semi & Oracle & Repetitions \\ \hline
\multicolumn{1}{l|}{Beck et al. \cite{beck2021effective}} & batch & 4 & 7 & \checkmark & - & - & - & - & - & - \\
\multicolumn{1}{l|}{Hu et al. \cite{hu2021towards}} & batch & 5 & 13 & \checkmark & \checkmark & - & - & - & - & 3 \\
\multicolumn{1}{l|}{Zhou et al. \cite{zhou2021towards}} & batch & 3 & 2 & \checkmark & \checkmark & - & - & - & \checkmark & 5 \\
\multicolumn{1}{l|}{Zhan et al. \cite{zhan2021comparative}} & sngl+batch & 35 & 18 & - & - & \checkmark & \checkmark & - & \checkmark & 10-100 \\
\multicolumn{1}{l|}{Munjal et al. \cite{munjal2022towards}} & batch & 2 & 8 & \checkmark & - & - & - & - & - & 3 \\
\multicolumn{1}{l|}{Li et al. \cite{li2022empirical}} & batch & 5 & 13 & \checkmark & - & - & - & \checkmark & - & - \\
\multicolumn{1}{l|}{Ji et al. \cite{ji2023randomness}} & batch & 3 & 8 & \checkmark & - & - & - & - & - & - \\
\multicolumn{1}{l|}{Luth et al. \cite{luth2024navigating}} & batch & 4 & 5 & \checkmark & - & - & - & \checkmark & - & 3 \\
\multicolumn{1}{l|}{\textbf{Ours}} & sngl+batch & 9(14) & 10 & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & 50 

@@

@@

\subsection*{Contributions}
\begin{enumerate}
	\item Evaluation of Active Learning algorithms on datasets from 5 different domains
	\item Two novel synthetic datasets that highlight principled shortcomings of existing AL algorithms
	\item Efficient and performant algorithm for an oracle that can be constructed greedily and does not rely on search
        \item Evidence that previous works might have not repeated their experiments often enough to provide conclusive results
\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Problem Description}
Given two spaces $\X, \Y$, $n=l+u$ data points with $l \in \mathbb{N}$ labeled examples $\mathcal{L} = \{(x_1, y_1),\ldots, (x_l,y_l)\}$, $u \in \mathbb{N}$ unlabeled examples $\mathcal{U} = \{x_{l+1},\ldots,x_{n}\}$, a model $\hat y: \X \to \Y$, a budget $\mathbb{N} \ni b \le u$ and an annotator $A: \mathcal{X} \to \mathcal{Y}$ that can label $x$ (we consider only hard labels in the one-hot format). 
We call $x \in \mathcal{X}$, $y \in \mathcal{Y}$ predictors and labels respectively where $(x,y)$ are drawn from an unknown distribution $\rho$. 
Find an acquisition function $\Omega: \U^{(i)},\LL^{(i)} \mapsto x^{(i)} \in \U^{(i)}$ that iteratively selects the next unlabeled point $x^{(i)}$ for labeling
\begin{align*}
	\LL^{(i+1)} &\gets \LL^{(i)} \cup \{\left(x^{(i)}, A(x^{(i)})\right)\} \\
	\U^{(i+1)} &\gets \U^{(i)} \setminus x^{(i)} %, i \in 1{:}B
\end{align*}
with $\U^{(0)} = \operatorname{seed}(\U, s)$ and $\LL^{(0)} = \{\left(\U^{(0)}, A(\U^{(0)})\right)\}$, where $\operatorname{seed}(\U, s)$ selects $s$ points per class for the initial labeled set. \\
So that the average expected loss $\ell: \mathcal{Y} \times \mathcal{Y} \to \mathbb{R}$ of a machine learning algorithm fitting $\hat y^{(i)}$ on the respective labeled set $\LL^{(i)}$ is minimal: 
$$\min \quad \frac{1}{B} \sum\limits_{i=0}^B \mathbb{E}_{(x,y) \sim \rho} \ell(y, \hat y^{(i)})$$

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Related Work}
Multiple benchmark suites have already been proposed for Active Learning:
The authors of \cite{beck2021effective}, \cite{munjal2022towards}, \cite{li2022empirical}, \cite{ji2023randomness} and \cite{luth2024navigating} focus exclusively on batch AL in the image domain.
While \cite{beck2021effective} discuss a new metric to measure AL performance, which they call ``Label Efficiency'' and provide experiments on many common configurations of data preparation, model training, and other hyperparameters, \cite{li2022empirical} focuses on combined approaches of AL and semi-supervised learning.
The authors of \cite{hu2021towards} study models that are trained on actively learned datasets in the image and text domain.
They test for several different properties of the models including robustness, response to compression techniques and final performance.
\cite{zhou2021towards} proposed an oracle algorithm for AL that uses Simulated Annealing search to approximate a solution for the optimal subset of labeled data.
Additionally, they study the generalization behavior of subsets of labeled data in the text an image domain.
The two closest related works to this benchmark are \cite{ji2023randomness} and \cite{luth2024navigating}, who also discussed the problems of evaluating AL algorithms under many forms of variance.
We largely adapt the proposed guidelines of \cite{ji2023randomness} and extend their work to multiple domains, batch sizes and comparisons. 
This work also pays attention to the so-called "pitfalls" of AL evaluation mentioned in \cite{luth2024navigating} with the exception of P1 - Controlling the data distribution of the tested datasets.
We leave this aspect for future versions of this benchmark. \\
Two preceding benchmarks have also proposed an oracle algorithm to find a near-optimal labeled set $\LL$, however, both algorithms rely on search and are for that reason considerably slower than our proposed greedy oracle.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Evaluation Protocol}\label{sec:evaluation}
Following \cite{zhou2021towards}, the quality of an AL algorithm is evaluated by an ``anytime protocol" that incorporates classification performance at every iteration, as opposed to evaluating final performance after the budget is exhausted.
We employ the normalized area under the accuracy curve (AUC):
\begin{equation}\label{eq:auc}
	\operatorname{AUC}(\D_\test, \hat y, B) := \frac{1}{B} \sum_{i=1}^{B} \operatorname{Acc}(\D_\test, \hat y^{(i)})
\end{equation}
The AUC incorporates performance in early stages (low budget) as well as capabilities to push the classifier in later stages (high budget).
AL algorithms have to perform well in both scenarios. 
\\ [1mm]
Since AUC is still influenced by the budget, we define a set of rules to set this hyperparameter upfront, so that we are not favoring a subset of algorithms by handcrafting a budget.
In this work, we choose the budget per dataset to be the first point at which one of 2 stopping conditions apply: (i) an algorithm (except Oracle) manages to reach 99\% of the full-dataset-performance (using the smallest query size) or (ii) the best algorithm (except oracle) did not improve the classifier's accuracy by at least 2\% in the last 20\% of iterations. \\ [1mm]
%\footnote{Since we only consider single-sample AL, increasing the budget comes at a high computational cost. Therefore we could not freely increase the budget above 2000.} 
To mimic the leave-one-out protocol for cross-validation, we will restart each experiment multiple times.
Each restart retains the train/test split (often given by the dataset itself), but introduces a new validation split that is sampled from the entire dataset (not just $\LL$).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Why we need 50 restarts}\label{sec:restarts}
To evaluate how many restarts are necessary to obtain conclusive and reproducible results in an AL experiment, we set the number of restarts of the margin sampling baseline for the Splice dataset to 100.
This allows us firstly, to obtain a very strong estimation of the "true" average performance of margin sampling on this dataset and secondly, to draw subsets from this pool of 100 runs.
Setting the size of our draws to $\alpha$ and sampling uniformly, we can approximate a cross-validation process with $\alpha$ restarts.
Each of these draws can be interpreted as a reported result in AL literature where the authors employed $\alpha$ restarts.
%We provide evidence in Fig. \ref{fig:restarts} that previous works might not have evaluated their experiments with a sufficient number of restarts.
Figure \ref{fig:restarts} shows the "true" mean performance of margin sampling (green) in relation to random sampling (black) and the oracle performance (red).
We display 5 random draws of size $\alpha$ in blue.
We can observe that even for a relatively high number of restarts the variance between the samples is extremely high, resulting in some performance curves being worse that random and some being significantly better.
When setting $\alpha = 50$ we observe all samples to converge close to the true mean performance. 
For that reason, we employ 50 restarts of every experiment in this work.
\begin{figure}
        \hspace{-20mm}
	\includegraphics[width=1.24\linewidth]{img/ablation_restarts}
	\caption{Random draws from a pool of 100 runs for margin sampling on the Splice dataset with different numbers of repetitions ($\alpha=\{5,10,50\}$). Green curves signify the true mean performance of all 100 runs, while the samples are blue. Even with 5 or 10 repetitions, we can observe that single draws for margin sampling display below-random performance (black), while the true mean should be above random.}
	\label{fig:restarts}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Seeding vs. Restarts}\label{sec:reproducibility}
Considering the high computational cost of 50 repetitions, another approach to ensure reproducibility would be to reduce the amount of variance in the experiment by keeping as many subsystems (weight initialization, data splits, etc.) as possible fixed with specialized seeding. \\
We describe a novel seeding strategy in Appendix \ref{app:seeding_strategy} that creates 3 separate Random Number Generators (RNG) based on 3 different seeds.
In summary, we introduce three different seeds: $s_\Omega$ for the AL algorithm, $s_\mathcal{D}$ for dataset splitting and mini-batch sampling, and $s_\theta$ for model initialization and sampling of dropout masks.
Unless stated otherwise, we will keep $s_\Omega$ fixed, while $s_\mathcal{D}$ and $s_\theta$ are incremented by 1 between restarts to introduce stochasticity into our framework.
While this seeding strategy is capable of controlling the amount variance in the experiment, previous works have noted that an actively sampled, labeled set does not generalize well between model architectures or even different initializations of the same model (\cite{zhou2021towards, lowell2018practical}), reducing its value in practice and providing a bad approximation of the quality of the AL algorithm.
Hence, we opt for letting the subsystems vary (by increasing $s_\mathcal{D}$ and $s_\theta$) and combine that with a high number of restarts to obtain a good average of the generalization performance of each AL algorithm. \\
Where a high number of restarts is computationally not feasible, we advise to additionally keep either $s_\mathcal{D}$ or $s_\theta$ (or both) fixed.

@@

\subsection{Datasets}\label{sec:datasets}
A detailed description of the preprocessing of each dataset can be found in Appendix \ref{app:hyperparameters}. \\ [1mm]
\textbf{Tabular:}
AL research conducted on tabular data is sparse (only \cite{ashdeep} from the considered baseline papers). 
We, therefore, introduce a set of tabular datasets that we selected according to the following criteria:
(i) They should be solvable by medium-sized models in under 1000 samples, (ii) the gap between most AL algorithms and random sampling should be significant (potential for AL is present) and (iii) the gap between the AL algorithms and our oracle should also be significant (research on these datasets can produce further lifts).
We use \textbf{Splice}, \textbf{DNA} and \textbf{USPS} from LibSVMTools \cite{libsvmtools}.\\
\textbf{Image:}
We use \textbf{FashionMNIST} \cite{xiao2017fashion} and \textbf{Cifar10} \cite{krizhevsky2009learning}, since both are widely used in AL literature.\\
\textbf{Text:}
We use \textbf{News Category} \cite{misra2022news} and \textbf{TopV2} \cite{chen-etal-2020-low-resource}.
Text datasets have seen less attention in AL research, but most of the papers that evaluate on text (\cite{hu2021towards}, \cite{zhou2021towards}) use at least one of these datasets.\\ [1mm]
We would like to point out that these datasets are selected for speed of computation (both in terms of number of features and necessary budget to solve the dataset). 
However, similar to our argumentation for picking smaller classifiers, we are solely focused on comparing different AL algorithms in this paper and do not aim to develop novel classification models on these datasets.
Our assumption is that a well-performing algorithm in our benchmark will also generalize well to larger real-world datasets, because we included multiple different data domains and classifier sizes in our experiments. \\ [1mm]
Adapting the experimental setting from \cite{hacohen2022active}, we offer all our datasets in the un-encoded (normal) setting as well as pre-encoded by a fixed embedding model that was trained by unsupervised contrastive learning. 
The text datasets are an exception to this, as they are only offered in their encoded form.
The pre-encoded datasets enable us to test single-sample algorithms on more complex datasets like Cifar10 and FashionMnist.
They also serve the purpose of investigating the interplay between self-supervised learning techniques and AL, as well as alleviating the cold-start problem described in \cite{luth2024navigating} as they require a way smaller seed set.
The classification model for every encoded dataset is a single linear layer with softmax activation.
The embedding model was trained with the SimCLR \cite{chen2020simple} algorithm adopting the protocol from \cite{hacohen2022active}. 
%For Cifar10 and FashionMnist we adapt the reported hyperparameters from \cite{hacohen2022active} and for the tabular datasets we use random search to optimize the hyperparameters.
To ensure that enough information from the data is encoded by our embedding model, the quality of embeddings during pretext training was measured after each epoch.
We attached a linear classification head to the encoder, fine-tuned it to the data and evaluated this classifier for test accuracy, mirroring our AL setup for embedded datasets. 
The checkpoints for each encoder model will be provided together with the framework. \\ [1mm]
\textbf{Synthetic Data}
Existing AL algorithms can broadly be categorized into two types, uncertainty \cite{wang2014new, gal2017deep} - and geometric-approaches \cite{sener2017active, hacohen2022active, ashdeep}. 
Both types have principled shortcomings in terms of the utilized information that makes them unsuitable for certain data distributions.
To test for these shortcomings, we created two synthetic datasets, namely "Honeypot" and "Diverging Sin", that are hard to solve for methods focused on the classifier's decision boundary or data clustering respectively. 
To avoid algorithms memorizing these datasets they are generated during the experiment, depending on $s_\D$. \\ [1mm]
\begin{figure}[H]

@@

	\includegraphics[width=0.4\linewidth]{img/diverging_sin.jpg}
	\includegraphics[width=0.4\linewidth]{img/honeypot.jpg}
	\caption{Our two synthetic datasets.}
        \label{fig:synthDataAppendix}

@@

Honeypot creates to two easy to distinguish clusters with 150 samples each and one overlapping "honeypot" that represents a noisy region of the dataset with potentially miss-labeled, miss-measured or generally adverse samples.
This honeypot contains 150 of each class, creating a balance of 50\% beneficial samples and 50\% adverse samples in the dataset.
The honeypot is located on the likely decision boundary of a classifier that is trained on the beneficial samples to maximize its negative impact on purely uncertainty based acquisition functions.
Diverging Sine samples the datapoints for each class from two diverging sinusoidal functions that are originating from the same y-intercept.
This creates a challenging region one the left hand side, where a lot of datapoints need to be sampled and an easy region on the right hand side, where very few datapoints are enough. 
The repeating nature of a sin function encourages diversity based acquisition functions to equally sample the entire length, drastically oversampling the right hand side of the dataset.
Each class has 500 datapoints. \\ [1mm]
